Autopolyploid establishment through polygenic adaptation

### 2025-01-14 doc/paper/paper.tex
paper update

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
\cite{barton2017}) for the inheritance of an additive quantitative trait in a
mixed-ploidy population consisting of diploid, triploid and autotetraploid
individuals producing haploid and diploid gametes.
We implement efficient simulation methods and use these to study the
quantitative genetics of mixed-ploidy populations and the establishment of
autotetraploids in a new habitat.
We show that in the absence of migration from the source population,
autotetraploids are more likely to found a successful population than diploids
under a very broad range of conditions, but that this is unlikely to
sufficiently counter the scarcity of tetraploid founders when the source is
predominantly diploid.
[...]
\end{abstract}

@@

Consider for instance a randomly mating diploid population.
Under the commonly accepted view that polyploids mostly emerge through the
union of unreduced gametes \citep{bretagnolle1995,herben2016,kreiner2017b}, a
new tetraploid individual originating by a chance encounter of two unreduced
diploid gametes (an event occurring at an appreciable rate; \cite{kreiner2017})
is highly unlikely to establish a stable tetraploid subpopulation, as most of
its gametes will end up in unfit hybrids of odd ploidy level (triploid block,
\cite{ramsey1998,kohler2010,brown2024}).
This negative frequency dependence effect is commonly referred to as
\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
It is well-appreciated that, due to MCE, in a large random mating mixed-ploidy
population dominated by diploids, the rate of unreduced gamete formation needs
to be extraordinarily high for tetraploids to establish (\cite{felber1997}, see
also \cref{s-sec:det}).

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of prezygotic isolation between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
localized dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
In particular, polyploids are thought to establish mainly in novel, unoccupied
habitats where they evade MCE (for instance at range edges, or after local
extinction due to environmental change).
However, such peripheral habitats are likely to present adaptive challenges to
establishment \citep{kawecki2008}, and if polyploids are able to colonize such
habitats at an appreciable rate, they must somehow be better adapted to local
conditions, or more able to adapt to those conditions despite inbreeding and
migration from the source population while the population is still small,
compared to diploids.
Indeed, when the source population is dominated by diploids, the probability
that a migrant individual is tetraploid will be very small ($O(u^2)$ if $u$ is
the probability that a diploid in the source population produces an unreduced
gamete in meiosis), so that tetraploids need a substantial advantage during
colonization if they are to establish before diploids do.

While there have been substantial modeling efforts aimed at understanding
autotetraploid establishment within diploid populations (e.g. \cite{levin1975,
felber1991, felber1997, rausch2005, oswald2011, clo2022c, griswold2021}), the
problem of polyploid establishment in a novel habitat which presents some adaptive
challenge remains largely unaddressed, despite its centrality to verbal
arguments about the establishment of polyploids in natural populations
\citep{kolar2017, vandepeer2021, clo2022d}.

Here we develop a model for the establishment of a mixed-ploidy population in a
novel, unoccupied habitat based on \cite{barton2018}.
In order to establish in the novel habitat, the population has to adapt to
local environmental conditions.
We assume fitness is determined by directional selection on a single polygenic
trait which can be interpreted as log fitness at low density in the new
habitat.
As in \cite{barton2018}, we assume the trait follows the infinitesimal
model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
infinitesimal model \citep{turelli2017}).
We extend the infinitesimal model, and the approach for exact simulation of
trait evolution under the infinitesimal model, to mixed-ploidy populations.
We then use simulations to study tetraploid establishment, both from single
migrants and under continuous migration from a predominantly diploid source
population, examining the effects of autopolyploid genetics,
maladaptive migration, selfing and assortative mating on the probability that
autotetraploids establish in the novel habitat.

@@

\begin{table}[t]
\caption{Glossary of the main notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N$ & total population size\\
$N_k$ & population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & selfing rate in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

Note that in a finite population, the segregation variance will decay over time
as the population becomes more inbred.
Indeed, Mendelian segregation generates less variation among the gametes
produced by individual $i$ when that individual is more inbred, as segregation
at homozygous loci does not generate any variation.
When $F_i$ is the inbreeding coefficient relative to some ancestral reference
population with gametic segregation variance $V$ (i.e. the probability that two
genes at a locus in individual $i$ sampled without replacement are identical by
descent), the gametic segregation variance of individual $i$ will be
$V_i = (1-F_i)V$.
This holds for both diploids and tetraploids (\cref{s-sec:tetinbred}, also
\cite{moody1993}).

@@

If we would somewhat naively assume that the allelic effects underlying an
additive quantitative trait are identical across ploidy levels, 
a tetraploid offspring from a cross between two diploids would yield, on average,
a trait value which is the sum of the parental trait values, with a
distribution that depends on the variance generated by the process producing
unreduced gametes.
This is not likely to reflect biological reality: tetraploids do not tend to
have, for instance, twice the size of their diploid progenitors on average.
Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
(HWLE) in tetraploids will be twice that of their diploid counterparts under
such assumptions, which is similarly unrealistic \citep{clo2022}.

In order to account for this, we introduce a scaling factor $\beta_k$,
accounting for the effects of polyploidization \textit{per se} on trait
expression in $k$-ploids.
To introduce and interpret this parameter, we consider an $L$-locus additive
model, with two alleles (0 and 1) at each locus.
For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
$i$.
We assume the trait value is determined by
\begin{equation}
  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
\end{equation}
Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
$k$-ploids.
The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
\begin{equation}
  \tilde{V}_{z,k} = k\sum_{i=1}^L a_{i,k}^2 p_iq_i = k\Vx_{k}
\end{equation}
where we refer to $\Vx_{k}$ as the variance associated with a haploid genome in
$k$-ploids at HWLE.
Note that we also have $\tilde{V}_{z,2} = 2\Vx_{2} = 2V$ \citep{barton2017},
where $V$ is the segregation variance in the diploid population.
If we now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
are as in diploids, but scaled by a factor $\beta_k$.
This implies that
\begin{equation}
  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
  = \frac{k\Vx_{k}}{2\Vx_{2}} = \frac{k}{2} \beta_{k}^2
\end{equation}
and hence also that $\Vx_{k} = \beta_k^2 \Vx_{2} = \beta_k^2 V$.
These relations will also hold in the infinitesimal limit where $a_{i,2}
\rightarrow 0$ as $L\rightarrow \infty$.

@@

We can extend the infinitesimal model to the mixed-ploidy case, assuming that the
gametic value, on the diploid trait scale, associated with a $k$-ploid gamete ($k
\in \{1,2\}$) from individual $i$ of ploidy level $c_i \in \{2,3,4\}$ is a
Gaussian random variable $Y_{i,k}$ with distribution
  $$Y_{i,k} \sim \Normal\left(\frac{k}{c_i}\frac{z_i}{\beta_{c_i}}, V_{i,k}\right)$$
where $V_{i,k}$ is the gametic segregation variance associated with the
production of a $k$-ploid gamete by individual $i$.
The trait value of an individual originating from the union of a $k$-ploid
gamete of individual $i$ and an $l$-ploid gamete from individual $j$ is then
  $$Z_{ij}^{kl} = \beta_{k+l}\left(Y_{i,k} + Y_{j,l}\right)$$
i.e., $Z_{ij}^{kl}$ is a Gaussian random variate with distribution
\begin{equation}
  Z_{ij}^{kl} \sim \Normal\left(
    \beta_{k+l} \left(
          \frac{k}{c_i}\frac{z_i}{\beta_{c_i}} 
        + \frac{l}{c_j}\frac{z_j}{\beta_{c_j}}\right), 
    \beta_{k+l}^2 (V_{i,k} + V_{j,l})\right)
   \label{eq:oneline}
\end{equation}

@@

\begin{table}[t]
\caption{Gametic segregation variance for haploid and diploid gametes produced
by the three cytotypes in the mixed-ploidy model. $F_i$ is the inbreeding
coefficient of individual $i$ (producing the gamete), whereas $\alpha_k$ is the
probability that a diploid gamete from a $k$-ploid individual receives two
copies of the same parental gene. Note that we have $\alpha_3 \le 1/4$ and
$\alpha_4 \le 1/6$.
} \label{tbl:segvar}
\centering
\small
\begin{tabularx}{\linewidth}{XXX}
\cline{1-3}
\textbf{cytotype}   & \textbf{haploid gamete variance} & \textbf{diploid gamete
variance}        \\ \cline{1-3} \\[-2.5ex]
diploid    & $\frac{1}{2}(1-F_i)V$    & $2\alpha_2(1-F_i)V$  \\ \\[-2.5ex]
triploid   & $\frac{2}{3}(1-F_i)V$     & $\frac{2}{3}(1 + 3\alpha_3)(1-F_i)V$ \\ \\[-2.5ex]
tetraploid & --                      & $(1+2\alpha_4)(1-F_i)V$             \\
\cline{1-3}
\end{tabularx}%
\end{table}

@@

Our model for the establishment of a population in an initially unoccupied
habitat is based on \cite{barton2018}.
We assume a large non-inbred `mainland' population at HWLE and cytotype
equilibrium, with $\Ex[z] = 0$ irrespective of the cytotype.
In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
distributed with mean $m$. 
The migrant individuals are assumed to be unrelated to the resident
individuals.
After migration in generation $t$, the $N(t) = N^\ast(t) + M(t)$ individuals
reproduce sexually, and the offspring thus produced survives until the next
generation with a probability determined by their trait value.
In the basic model, random selfing is allowed (but see below for a model with
self-incompatibility).
We assume the trait is under directional selection, with fitness is $w(z) =
e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
selection and $\theta$ is the trait value for which the growth rate of the
population becomes positive.

@@

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
\caption{
The infinitesimal model in autotetraploids.
Comparisons are shown for the decay of the genetic variance ($V_z$) due to
inbreeding in exact simulations of the infinitesimal model in autotetraploids
against individual-based simulations of autotetraploid populations with $L$
unlinked additive loci determining the quantitative trait. 
(A) Simulations of a model without double reduction ($\alpha_4=0$).
(B) Simulations of a model with maximal double reduction ($\alpha_4=1/6$) (for
all loci in the finite $L$ simulations).
We show window-smoothed values for visual clarity, with observed genetic
variances averaged in windows of 20 generations every 10 generations.
The black line marks $e^{-t/4N}$. 
We assume $N=250$ and $V_z(0) = 1$.
\label{fig:vztet}}
\end{figure}

We evaluate the accuracy of the autotetraploid infinitesimal model as an
approximation to the evolution of a quantitative trait determined by $L$
additive loci.
We find that the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance when the number of loci
is sufficiently large ($L \ge 100$, say, \cref{fig:vztet,s-fig:fininf}).
Furthermore, we confirm that, in the absence of double reduction, the decay in
genetic variance due to inbreeding after a time $t$ is well-predicted by
$e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
As predicted, double reduction leads to an immediate increase in genetic
variance, but leads to accelerated inbreeding, causing faster loss of genetic
variance in the long-term (\cref{fig:vztet,s-fig:fininf}).
As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
double reduction cannot, however, be described by a single long-term effective
population size.

>>>>

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
\cite{barton2017}) for the inheritance of an additive quantitative trait in a
mixed-ploidy population consisting of diploid, triploid and autotetraploid
individuals producing haploid and diploid gametes.
We implement efficient simulation methods and use these to study the
quantitative genetics of mixed-ploidy populations and the establishment of
autotetraploids in a new habitat.
We show that for rare migration from the source population, autotetraploids are
more likely to found a successful population than diploids under a very
broad range of conditions, but that this is unlikely to sufficiently
counter the scarcity of tetraploid founders when the source is
predominantly diploid.

\end{abstract}

@@

Consider for instance a randomly mating diploid population.
Under the commonly accepted view that new polyploid plants are mostly formed
through the union of unreduced gametes
\citep{bretagnolle1995,herben2016,kreiner2017b}, a new tetraploid individual
originating by a chance encounter of two unreduced diploid gametes (an event
occurring at an appreciable rate; \cite{kreiner2017}) is highly unlikely to
establish a stable tetraploid subpopulation, as most of its gametes will end up
in unfit hybrids of odd ploidy level (a phenomenon referred to as `triploid
block', see \cite{ramsey1998,kohler2010,brown2024}).
This negative frequency dependence effect is commonly referred to as
\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
It is well-appreciated that, as a consequence of MCE, the rate of unreduced
gamete formation needs to be extraordinarily high for tetraploids to establish
in a large random mating initially diploid population (\cite{felber1997}, see
also \cref{s-sec:det}).

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of \textit{prezygotic isolation} between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
localized dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
In particular, polyploids are thought to establish mainly in novel, unoccupied
habitats where they evade MCE (for instance at range edges, or after local
extinction due to environmental change) \citep{griswold2021}.
However, such peripheral habitats are likely to present adaptive challenges to
establishment \citep{kawecki2008adaptation}, and if polyploids are able to
colonize such habitats at an appreciable rate, they must somehow be better
adapted to local conditions, or more able to adapt to those conditions despite
inbreeding and maladaptive migration, compared to diploids.
Indeed, when the source population is dominated by diploids, the probability
that a migrant individual is tetraploid will be very small ($O(u^2)$ if $u$ is
the probability that a diploid in the source population produces an unreduced
gamete in meiosis), so that tetraploids need a substantial advantage during
colonization if they are to establish before diploids do.

While there have been substantial modeling efforts aimed at understanding
autotetraploid establishment within diploid populations \citep{levin1975,
felber1991, felber1997, rausch2005, oswald2011, clo2022c}, the problem of
polyploid establishment in peripheral habitats remains largely unaddressed
(but see \cite{griswold2021}),
despite its centrality to verbal arguments about the establishment of
polyploids in natural populations \citep{kolar2017, vandepeer2021, clo2022d}.

Here we develop a rather general model for the establishment of a mixed-ploidy
population in a novel, unoccupied habitat based on \cite{barton2018}.
In order to establish in the novel habitat, the population has to adapt to
local environmental conditions.
We assume fitness is determined by directional selection on a single polygenic
trait which can be interpreted as log fitness at low density in the new
habitat.
As in \cite{barton2018}, we assume the trait follows the infinitesimal
model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
infinitesimal model \citep{turelli2017}).
We extend the infinitesimal model, and the approach for exact simulation of
trait evolution under the infinitesimal model, to mixed-ploidy populations.
This is a first contribution of the present paper.
We then use simulations to study tetraploid establishment, both from single
migrants and under continuous migration from a predominantly diploid source
population, examining the effects of autopolyploid genetics,
maladaptive migration, selfing and assortative mating on the probability that
autotetraploids establish in the novel habitat.

@@

\begin{table}[t]
\caption{Glossary of the notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N$ & total population size\\
$N_k$ & population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & rate of self-fertilization in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

In a finite population, the segregation variance will decay over time
as the population becomes more inbred.
Indeed, Mendelian segregation generates less variation among the gametes
produced by individual $i$ when that individual is more inbred, as segregation
at homozygous loci does not generate any variation.
When $F_i$ is the inbreeding coefficient relative to some ancestral reference
population with gametic segregation variance $V$ (i.e. the probability that two
genes at a locus in individual $i$ sampled without replacement are identical by
descent), the gametic segregation variance of individual $i$ will be
$V_i = (1-F_i)V$.
This holds for both diploids and tetraploids (\cref{s-sec:tetinbred}, also
\cite{moody1993}).

@@

If we would somewhat naively assume that the allelic effects underlying an
additive quantitative trait are identical across ploidy levels, 
a tetraploid offspring from a cross between two diploids would yield, on average,
a trait value which is the sum of the parental trait values, with a
distribution that depends on the variance generated by the process producing
unreduced gametes.
This is not likely to reflect biological reality: tetraploids do not tend to
have, for instance, twice the size of their diploid progenitors on average.
Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
(HWLE) in a large non-inbred tetraploid population will be twice that of their
diploid counterparts under such assumptions, which is similarly unrealistic
\citep{clo2022}.

In order to account for this, we introduce a scaling factor $\beta_k$,
accounting for the effects of polyploidization \textit{per se} on trait
expression in $k$-ploids.
To introduce and interpret this parameter, we consider an $L$-locus additive
model, with two alleles (0 and 1) at each locus.
For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
$i$.
We assume the trait value is determined by
\begin{equation}
  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
\end{equation}
Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
$k$-ploids.
The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
\begin{equation}
  \tilde{V}_{z,k} = k\sum_{i=1}^L a_{i,k}^2 p_iq_i = k\Vx_{k}
\end{equation}
where we refer to $\Vx_{k}$ as the variance associated with a haploid genome in
$k$-ploids at HWLE.
If we now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
are as in diploids, but scaled by a factor $\beta_k$, and assume equal allele
frequencies in the different cytotypes, 
we find that
\begin{equation}
  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
  = \frac{k\Vx_{k}}{2\Vx_{2}} = \frac{k}{2} \beta_{k}^2
\end{equation}
Note that by definition $\beta_2=1$.
Under the infinitesimal model (where $a_{i,2} \rightarrow 0$ as $L\rightarrow
\infty$), we have $\tilde{V}_{z,2} = 2\Vx_{2} = 2V$ \citep{barton2017}, where
$V$ is the segregation variance in the diploid population.
Hence, in the infinitesimal limit we have $\Vx_{k} = \beta_k^2 \Vx_{2} =
\beta_k^2 V$.

@@

We can extend the infinitesimal model to the mixed-ploidy case, assuming that the
gametic value, on the diploid trait scale, associated with a $k$-ploid gamete ($k
\in \{1,2\}$) from individual $i$ of ploidy level $c_i \in \{2,3,4\}$ is a
Gaussian random variable $Y_{i,k}$ with distribution
\begin{equation}
    Y_{i,k} \sim \Normal\left(\frac{k}{c_i}\frac{z_i}{\beta_{c_i}}, V_{i,k}\right)
    \label{eq:gamete}
\end{equation}
where $V_{i,k}$ is the gametic segregation variance associated with the
production of a $k$-ploid gamete by individual $i$ (see below).
The trait value of an individual originating from the union of a $k$-ploid
gamete of individual $i$ and an $l$-ploid gamete from individual $j$ is then
  $$Z_{ij}^{kl} = \beta_{k+l}\left(Y_{i,k} + Y_{j,l}\right)$$
i.e., $Z_{ij}^{kl}$ is a Gaussian random variate with distribution
\begin{equation}
  Z_{ij}^{kl} \sim \Normal\left(
    \beta_{k+l} \left(
          \frac{k}{c_i}\frac{z_i}{\beta_{c_i}} 
        + \frac{l}{c_j}\frac{z_j}{\beta_{c_j}}\right), 
    \beta_{k+l}^2 (V_{i,k} + V_{j,l})\right)
   \label{eq:oneline}
\end{equation}

@@

\begin{table}[t]
\caption{Gametic segregation variance for haploid and diploid gametes produced
by the three cytotypes in the mixed-ploidy model. $F_i$ is the inbreeding
coefficient of individual $i$ (producing the gamete), whereas $\alpha_k$ is the
probability that a diploid gamete from a $k$-ploid individual receives two
copies of the same parental gene. We assume $\alpha_3 \le 1/4$ and
    $\alpha_4 \le 1/6$ (see \cref{s-sec:segvar}).
} \label{tbl:segvar}
\centering
\small
\begin{tabularx}{\linewidth}{XXX}
\cline{1-3}
\textbf{cytotype}   & \textbf{haploid gamete variance} & \textbf{diploid gamete
variance}        \\ \cline{1-3} \\[-2.5ex]
diploid    & $\frac{1}{2}(1-F_i)V$    & $2\alpha_2(1-F_i)V$  \\ \\[-2.5ex]
triploid   & $\frac{2}{3}(1-F_i)V$     & $\frac{2}{3}(1 + 3\alpha_3)(1-F_i)V$ \\ \\[-2.5ex]
tetraploid & --                      & $(1+2\alpha_4)(1-F_i)V$             \\
\cline{1-3}
\end{tabularx}%
\end{table}

Writing $\bar{z}_2$ for the mean trait value in the diploid subpopulation,
\cref{eq:oneline} implies that a tetraploid offspring from a random diploid
parental pair has an expected trait value equal to $\bar{z}_4 =
2\beta_4\bar{z}_2$.
This hence implies that $|\bar{z}_4| \ge |\bar{z}_2|$, with equality only when
$\beta_4=1/2$ or $\bar{z}_2 = \bar{z}_4 = 0$. 
In other words, when $\beta_4 > 1/2$, we would have for all but $\bar{z}_2 = 0$
that newly formed tetraploids have more extreme phenotypes on average than
their diploid parents.
In our establishment model (see below), we shall therefore always consider 
the case where $\bar{z}_2=0$ in the source population, and think of the trait
value modeled as the deviation from the mean phenotype in the source
population.

The property that polyploid phenotypes are more extreme on average makes sense
if we consider the underlying genetic model: if we assume the source population
is at an equilibrium between mutation, drift and stabilizing selection,
segregating genetic variants that affect the trait will be equally likely to
have positive or negative allelic effects \citep{hayward2022}.
At equilibrium, any diploid individual with trait value $z - \bar{z}_2 > 0$
then carries an excess of variants with positive allelic effects on the trait,
and this excess should be exaggerated in its polyploid offspring, which carry
\textit{twice} the excess of their parent on average, unless allelic effects are
scaled exactly by one half (i.e. $\beta_4 = 1/2$). 
In other words, the mixed-ploidy model is not `coordinate free' as the basic
infinitesimal model is: \cref{eq:oneline} only makes sense when the trait
values that are modeled correspond to deviations from the mean values
associated with an underlying equilibrium state.

@@

Our model for the establishment of a population in an initially unoccupied
habitat is based on \cite{barton2018}.
We assume a large non-inbred `mainland' mixed-ploidy population at HWLE and
cytotype equilibrium, with $\Ex[z] = 0$ irrespective of the cytotype.
The equilibrium trait value distribution for the different cytotypes on the
mainland is complicated in general, but a very accurate approximation (which we
use throughout) is readily obtained for the case where $u$ is small (see
\cref{s-sec:mixeq}).

In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
distributed with mean $m$. 
The migrant individuals are assumed to be unrelated to the resident
individuals.
After migration in generation $t$, the $N(t) = N^\ast(t) + M(t)$ individuals
reproduce sexually, and the offspring thus produced survives until the next
generation with a probability determined by their trait value.
In the basic model, random selfing is allowed (but see below for a model with
self-incompatibility).
We assume the trait is under directional selection, with fitness $w(z) =
e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
selection and $\theta$ is the trait value for which the growth rate of the
population becomes positive.
Throughout the paper, we define establishment as reaching a population of size
100.

@@

The results section is organized as follows: first we verify the correctness
and accuracy of the infinitesimal model for mixed-ploidy populations by
comparison against simulations of a model with finitely many loci.
Next we assess how autotetraploid genetics and different assumptions on the
genetic variance of neotetraploids affect the probability of tetraploid
establishment in a marginal habitat relative to diploid establishment, starting
from a single maladapted migrant individual.
We then study how continuous migration from a predominantly diploid maladapted
source population affects the relative establishment probability, assessing the
impact of migration load and MCE.
Finally, we consider the impact of prezygotic isolation mechanisms on
tetraploid establishment in the marginal habitat.
Throughout, we assume $\beta_3 = \sqrt{2/3}, \beta_4=\sqrt{1/2}$
and $\alpha_2 = \alpha_3 = \alpha_4 = 0$, unless stated otherwise.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/tetfininf.pdf}
\caption{
The infinitesimal model in autotetraploids.
Comparisons are shown for the decay of the genetic variance ($V_z$) due to
inbreeding in exact simulations of the infinitesimal model in autotetraploids
against individual-based simulations of autotetraploid populations with $L$
unlinked additive loci determining the quantitative trait. 
(A) Simulations of a model without double reduction ($\alpha_4=0$).
(B) Simulations of a model with maximal double reduction ($\alpha_4=1/6$) (for
all loci in the finite $L$ simulations).
We show window-smoothed values for visual clarity, with observed genetic
variances averaged in windows of 20 generations every 10 generations.
The black line marks $e^{-t/4N}$. 
We assume $N=250$ and $V_z(0) = 1$.
\label{fig:vztet}}
\end{figure}

We evaluate the accuracy of the autotetraploid infinitesimal model as an
approximation to the evolution of a quantitative trait determined by $L$
additive loci.
We find that the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance when the number of loci
is sufficiently large ($L \ge 100$, say, \cref{fig:vztet,s-fig:fininf}).
Furthermore, we confirm that, in the absence of double reduction, the decay in
genetic variance due to inbreeding after a time $t$ is well-predicted by
$e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
As predicted, double reduction leads to an immediate increase in genetic
variance (as it increases the segregation variance), but leads to accelerated
inbreeding, causing faster loss of genetic variance in the long-term
(\cref{fig:vztet,s-fig:fininf}).
As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
double reduction cannot, however, be described by a single long-term effective
population size.

Simulations for the mixed-ploidy model further confirm the correctness of our
infinitesimal approximation (\cref{s-fig:vz}), yielding accurate predictions
for the evolution of the phenotypic variance across ploidy levels and for
different assumptions regarding the scaling of allelic effects ($\beta$
parameters) and unreduced gamete formation ($\alpha$ parameters).
